---
title: "community-figures"
author: "K.L. Beck"
date: "6/16/2020"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(dplyr)
library(ggplot2)
library(gridExtra)
library(RColorBrewer)
library(robCompositions)
library(reshape)
library(stringr)
library(vegan)

options(scipen = 10)
```

## TODOs:
- Update alpha data div to sort by date and color by category
- split out supported plot as 1% going to other and viz stack and beta div
- split out supported plot as top 25 or quartile and viz stack and beta div
- beta div -- can this be colored by alpha diversity?
- Ban will work on updating ordination plots, plus correlation plots
- Niina will run hulk
- TODO generate a table on just controls - keffir is a positive control


## Visualizations for Anomaly Detection Prelim Samples

```{r loadData}

# load microbe tables
microbePath = "~/Box Sync/Sequencing the Food Supply Chain Consortium - Cornell + Penn Project/Shared-Data/Milk66samples_NewFiltering_microbe_genus_DS10828562.txt"
microbes_raw = read.csv(microbePath, header = TRUE, stringsAsFactors = FALSE, sep = "\t")

microbes_reads = microbes_raw
colnames(microbes_reads) = str_remove(colnames(microbes_reads), "Cumul.reads_dt")


# order columns by sorted sample name
microbes_reads = cbind(microbes_raw$Name, microbes_reads[,order(colnames(microbes_reads))])
microbes_reads$Name = NULL
colnames(microbes_reads)[1] = "Genus_Name"

# uncomment this for read count export to tabular data
# write.csv(microbes_reads, "~/read_count.txt", quote = FALSE, row.names = FALSE)

```

```{r computeRPM}

summaryFile = "~/Box Sync/Sequencing the Food Supply Chain Consortium - Cornell + Penn Project/Shared-Data/Milk66samples_NewFiltering_ReadCounts_forRPM.txt"
macros_raw = read.table(summaryFile, sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# reorder for easier merging below
macros = arrange(macros_raw, Display.TAG)


microbes_psct = microbes_reads

# additionally pseudo-count raw reads assigned to compute adjusted RPM
scale_const = 1000000
microbes_psct[, -1] = microbes_psct[, -1] + 1        # add one, removing all zeros

microbes_psct[,2:ncol(microbes_psct)] = sapply(2:ncol(microbes_psct), function(n) {
	microbes_psct[,n] / macros$Kraken.UNCLASSIFIED[n-1] * scale_const}
	)

# Export all microbes in RPM
# write.table(microbes_psct, file = "~/Box Sync/Sequencing the Food Supply Chain Consortium - Cornell + Penn Project/Shared-Data/milk-all-samples-new-filtering-RPM-all-microbes_from_DS10828562.txt", sep = "\t", quote = FALSE, row.names = FALSE)

```

```{r subsetGenusTable}

control_samples = c("NEG_DNAextr*", "Kefir*", "SequencingBlank*")
microbes = dplyr::select(microbes_psct, -matches(control_samples)) 
# this matches the Kraken datafile Niina added to github (note no control, keffir, etc)
colnames(microbes) = str_remove(colnames(microbes), "_ds[0-9]+")


rpm_threshold = 0.1

# microbes present above rpm threshold in  **all** samples
consensus_index = sapply(1:nrow(microbes), function(n) all(microbes[n,2:ncol(microbes)] > rpm_threshold))
consensus_microbes = microbes[consensus_index==TRUE, ]

consensus_microbes_melted = melt(consensus_microbes)

# microbes present above rpm threshold in  **any** samples
supported_index = sapply(1:nrow(microbes), function(n) any(microbes[n,2:ncol(microbes)] > rpm_threshold))
supported_microbes = microbes[supported_index==TRUE, ]

supported_microbes_melted = melt(supported_microbes)
# write.table(supported_microbes, file = "~/Box Sync/Sequencing the Food Supply Chain Consortium - Cornell + Penn Project/Shared-Data/milk-real-samples-new-filtering-RPM-supported-microbes-only_from_DS10828562.txt", sep = "\t", quote = FALSE, row.names = FALSE)


# extract dates and sample category from sample names

treatment_names = sapply(1:ncol(supported_microbes), function(n) 
	c(paste(str_split(string = colnames(supported_microbes), pattern = "_")[[n]][1:2], collapse = "_"))
	)

sample_dates = sapply(1:ncol(supported_microbes), function(n) 
	str_split(string = colnames(supported_microbes), pattern = "_")[[n]][3]
	)

sample_details = data.frame("full_sample_name" = colnames(supported_microbes)[-1],
							"sample_dates"     = sample_dates[-1],
							"treatment_names"  = treatment_names[-1]
							)

```




## Microbial Community
```{r viz}

ggplot(consensus_microbes_melted, aes(x = variable, y = value, fill = Genus_Name)) +
	geom_bar(stat = "identity") + 
	ylab("Relative abundance") + xlab("Sample") +
	ggtitle(label = "Consensus genera microbial composition") +
	theme(axis.text.x = element_text(angle = 90, hjust = 1))


ggplot(supported_microbes_melted, aes(x = variable, y = value, fill = Genus_Name)) +
	geom_bar(stat = "identity") + 
	ylab("Relative abundance") + xlab("Sample") +
	ggtitle(label = "Supported genera microbial composition") +
	theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Ecological metrics - alpha diversity
```{r alpha_div}

alpha_div = data.frame("sample" = colnames(supported_microbes)[-1],
											 "nMicrobes" = sapply(2:ncol(supported_microbes), 
			 																function(x) 
																				sum(supported_microbes[ , x] >= rpm_threshold))
)

alpha_div$sample = factor(alpha_div$sample, levels = alpha_div$sample)

median_baseline_div = median(alpha_div$nMicrobes[which(grepl("Baseline*", alpha_div$sample))])

ggplot(alpha_div, aes(x=sample, y=nMicrobes)) +
	geom_bar(stat="identity") +
	theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
	geom_hline(yintercept = median_baseline_div)


```

## Ecological metrics - beta diversity

```{r beta_div}
rownames(supported_microbes) = supported_microbes$Genus_Name
ad = aDist(t(supported_microbes[,-1]))
hc_complete = hclust(ad, method = "complete")
hc_average  = hclust(ad, method = "average")
# clustering method could be complete (default) or any other 
# http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.876.3979&rep=rep1&type=pdf

# par(mfrow=c(2,1))
plot(hc_complete, main="Aitchison distance- Complete clustering", ylab="Distance", xlab="Sample Identifier", hang = -1)
plot(hc_average,  main="Aitchison distance- Average clustering", ylab="Distance", xlab="Sample Identifier", hang = -1)
# hang = -1 recommended by bala

# as recommended by Gloor froniers in microbiology supplement 2016
# https://github.com/ggloor/Frontiers_2017/blob/master/Frontiers_supplement.Rmd
# coloring from https://rpubs.com/gaston/dendrograms
# make the dendrogram
hc <- as.dendrogram(hclust(ad, method="ward.D2"))
hcd <- hclust(ad, method="ward.D2")
# function to get color labels
colLab <- function(n) {
    if (is.leaf(n)) {
        a <- attributes(n)
        #labCol <- labelColors[clusMember[which(names(clusMember) == a$label)]]
        # MFMB-04, 20, 38 are all host contaminated
        # MFMB- <80 are the first supplier
        # MFMB- >=80 are the second supplier
        labCol <- if (grepl("Outside*", a$label) == TRUE ) "red" else if (grepl("TxAntibiotic", a$label) == TRUE ) "#339933" else "#6699FF"
        attr(n, "nodePar") <- c(a$nodePar, lab.col = labCol)
    }
    n
}
# using dendrapply
clusDendro = dendrapply(hc, colLab)
par(mar = c(11, 2, 2, 2))
plot(clusDendro, main="Beta diversity clustering by sample (Aitchison distance)", ylab="Aitchison distance")
legend("topright", 
     legend = c("Baseline" , "Tx Antibiotic" , "Outside Farm"), 
     col = c("#6699FF", "#339933", "red"), 
     pch = 20, bty = "n",  pt.cex = 1.5, cex = 1, 
     text.col = c("#6699FF", "#339933", "red"), horiz = FALSE, inset = c(0.05, 0.03))
```
