---
title: "explore-decontam"
author: "K.L. Beck"
date: "4/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(stringr)
library(decontam)
```

## Objective:  
Explore `decontam` package for use in removing false positives due to lab preparation

```{r loadData}

# load microbe tables
microbePath = "~/Box Sync/Sequencing the Food Supply Chain Consortium - Cornell + Penn Project/Shared-Data/Milk66samples_NewFiltering_microbe_genus_DS10828562.txt"
microbes_raw = read.csv(microbePath, header = TRUE, stringsAsFactors = FALSE, sep = "\t")

microbes_reads = microbes_raw
colnames(microbes_reads) = str_remove(colnames(microbes_reads), "Cumul.reads_dt")


# order columns by sorted sample name
microbes_reads = cbind(microbes_raw$Name, microbes_reads[,order(colnames(microbes_reads))])
microbes_reads$Name = NULL
colnames(microbes_reads)[1] = "Genus_Name"

# load macros for RPM calculation
summaryFile = "~/Box Sync/Sequencing the Food Supply Chain Consortium - Cornell + Penn Project/Shared-Data/Milk66samples_NewFiltering_ReadCounts_forRPM.txt"
macros_raw = read.table(summaryFile, sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# reorder for easier merging below
macros = arrange(macros_raw, Display.TAG)


microbes_psct = microbes_reads

# additionally pseudo-count raw reads assigned to compute adjusted RPM
scale_const = 1000000
microbes_psct[, -1] = microbes_psct[, -1] + 1        # add one, removing all zeros

microbes_psct[,2:ncol(microbes_psct)] = sapply(2:ncol(microbes_psct), function(n) {
	microbes_psct[,n] / macros$Kraken.UNCLASSIFIED[n-1] * scale_const}
	)

```


### Explore `decontam` Package on RPM Table  
See tutotial details in package [vignette](https://benjjneb.github.io/decontam/vignettes/decontam_intro.html)

```{r exploreDecontam}
# ----- eval decontam w rounded RPM values ----- #
# decontam input: rows = samples, organisms = columns
drop_samples = c("Kefir*", "SequencingBlank*")
microbes_w_controls = dplyr::select(microbes_psct, -matches(drop_samples)) 
# note ^ no RPM threshold applied here. this can be applied after decontamination

genera_labels = microbes_w_controls$Genus_Name
microbes_w_controls_t = t(data.matrix(microbes_w_controls[ ,-1]))
colnames(microbes_w_controls_t) = genera_labels

# round RPM values for integer requirement
microbes_w_controls_t_ceiling = ceiling(microbes_w_controls_t)
microbes_w_controls_t_floor = floor(microbes_w_controls_t)

# extraction controls
neg_labels = grepl("Neg", row.names(microbes_w_controls_t))

contaminants_ceil = isContaminant(microbes_w_controls_t_ceiling, neg = neg_labels, threshold = 0.5, normalize = FALSE)
contaminants_floor = isContaminant(microbes_w_controls_t_floor, neg = neg_labels, threshold = 0.5, normalize = FALSE)
# prevalence method is for negative DNA sequence controls (opposed to DNA concentration values)
table(contaminants_ceil$contaminant)
table(contaminants_floor$contaminant)
```

**Notes:**  
- normalize param does not change our results either way  
- decontam does have an internal normalization mechanism but this sums everything to 1, which breaks the conventions of CoDA and should not be use  
- question asked here: https://github.com/benjjneb/decontam/issues/45  
- best approximation is to convert our floats to a rounded integer, note here: https://github.com/benjjneb/decontam/issues/45  
- the difference between contaminant count based on taking the floor or ceiling of the RPM value is highly suspect and likely not a valid result for removal of contaminants  

### Explore `decontam` Package on Raw Reads
```{r decontamRawReads}
# ----- eval decontam w raw reads ----- #
# decontam input: rows = samples, organisms = columns
drop_samples = c("Kefir*", "SequencingBlank*")
microbes_w_controls_r = dplyr::select(microbes_reads, -matches(drop_samples)) 

genera_labels_r = microbes_w_controls_r$Genus_Name
microbes_w_controls_r_t = t(data.matrix(microbes_w_controls_r[ ,-1]))
colnames(microbes_w_controls_r_t) = genera_labels_r

# extraction controls
neg_labels = grepl("Neg", row.names(microbes_w_controls_r_t))

contaminants_r = isContaminant(microbes_w_controls_r_t, neg = neg_labels, threshold = 0.5, normalize = TRUE)
# prevalence method is for negative DNA sequence controls (opposed to DNA concentration values)

table(contaminants_r$contaminant)

```
**Notes:**  
- normalization only changes frequency, it does not change number of on contaminants or p-values  
- performing decontamination on raw reads is preferred. the results are more robust against the package's internal normalization and do not artifically force rounding of floating point numbers  
- decision: move forward with decontaminating at the raw reads point  
